Mixing Time Estimation in Reversible Markov Chains from a Single Sample Path

### 2015-06-09 arxiv-v1/intro.tex
remove comments

%!TEX root =  matrix-est.tex

%Performance guarantees play a key role in the analysis and design of machine learning algorithms.

This work
tackles the challenge of constructing
fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The \emph{mixing time} $\tmix$ of the Markov chain is the
number of time steps required
%time that it
%takes
for the chain to
be within a fixed threshold of
%(approximately) reach
its stationary
distribution:
\todoa{slightly changed wording}
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that
succeeds with probability $1-\delta$
in trapping
the value of the
mixing time $\tmix$.
\todoa{changed wording}

This problem is motivated by the numerous scientific applications and
machine learning tasks in which
the quantity of interest is
%one is interested in estimating
the
mean $\vpi(f) = \sum_i \pi_i f(i)$
for some function $f$ of the states of
a Markov chain.
\todoa{changed wording} \todoc{I am not sure that the new wording is better. It
fails to express that there is an estimation problem. "Quantity of interest"
is way to vague for me.} 
This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
paradigm~\cite{liu2001monte}, but the problem also arises in
performance prediction involving time-correlated data, as is common in
reinforcement learning~\cite{sutton98}.
Observable bounds on mixing times are useful in the design and
diagnostics of these methods; they yield effective approaches to
assessing the estimation quality, even when \emph{a priori} knowledge
of the mixing time or correlation structure is unavailable.

\paragraph{Main results.}
We develop the first procedure for constructing non-trivial
and fully empirical confidence intervals for Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with absolute
spectral gap $\gap$ and stationary distribution minorized by $\pimin$. 
As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
\begin{equation}
  \label{eq:mixing-time-bound}
  \Parens{\frac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \frac1{\gap} \ln \frac4{\pimin}
  ,
\end{equation}
%$(\frac{1}\gap  - 1)\ln(2)  \le \tmix \le \frac{1}\gap \ln \frac4\pimin$,
and hence it suffices to estimate $\gap$ and $\pimin$. 
%these quantities $\gap$ and $\pimin$ are well-known to govern
%mixing time.
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{sec:rates-upper}, we give a point-estimator for $\gap$,
    and prove in \cref{thm:err} that it achieves multiplicative
    accuracy from \emph{a single sample path} of length
    $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide a point-estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    This establishes the feasibility of \emph{estimating} the mixing
    time in this setting.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.
    We also discuss the importance of reversibility, and some possible
    extensions to non-reversible chains.

  \item
    In \cref{sec:rates-lower}, we show that $n = \Omega\parens{ (d\log
    d)/\gap + 1/\pimin}$ observations are necessary for any procedure
    to guarantee constant multiplicative accuracy in estimating $\gap$
    (\cref{thm:lb-pimin,thm:lb-gap}).
    Essentially, \emph{every} state may need to be visited about
    $\log(d)/\gap$ times, on average, before an accurate estimate of
    the mixing time can be provided, regardless of the estimation
    procedure used.

  \item
    In \cref{sec:empirical}, we develop a procedure for computing
    \emph{valid empirical confidence intervals} for $\pimin$ and
    $\gap$ (\cref{alg:empest}).
    The procedure uses just a single sample path and does not require
    any prior knowledge of the chain's parameters, such as $\pimin$ or
    $\gap$.
    We quantify the rate at which the width of the confidence
    intervals shrink as a function of the length of the sample path
    (\cref{thm:empirical}).
    We also explain how to combine the empirical confidence intervals
    from \cref{alg:empest} with the non-empirical bounds from
    \cref{thm:err} to produce valid empirical confidence intervals.
    We prove in \cref{thm:combined} that the width of these new
    intervals converge to zero asymptotically at least as fast as
    those from either \cref{thm:err} and \cref{thm:empirical}.

%    which is
%    $     O\Parens{
%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
%      +
%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
%    }.$

%    \todoc{please add rate}

@@

Although these asymptotic results help us understand the limiting
behavior of the sample mean over a Markov chain, they say little about
the finite-time non-asymptotic behavior, which is often needed for the
prudent evaluation of a method or even its algorithmic design
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}. %BaHaVa10:activelearning,ShaVo08:conformalpred
To address this need, numerous works have developed Chernoff-type
bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
valuable tools for non-asymptotic probabilistic
analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,
%meroi2012-jap,
paulin15}.
These probability bounds are larger than corresponding bounds for
independent and identically distributed (iid) data due to the temporal
dependence; intuitively, for the Markov chain to yield a fresh draw
$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
$\Theta(\tmix)$ time steps.
Note that the bounds generally depend on distribution-specific
properties of the Markov chain (e.g., $\vP$, $\tmix$, $\gap$), which are often
unknown \emph{a priori} in practice.
Consequently, much effort has been put towards estimating these
unknown quantities, especially in the context of MCMC diagnostics, in
order to provide data-dependent assessments of estimation
accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
However, these approaches generally only provide asymptotic
guarantees, and hence fall short of our goal of empirical bounds that
are valid with any finite-length sample path.

@@

It is possible to eliminate many of the difficulties presented above
%with additional access to the Markov chain.
when allowed more flexible access
to the Markov chain.
For example, given a sampling oracle that generates
independent transitions from any given state (akin to a ``reset''
device), the mixing time becomes an efficiently testable property in
the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
On the other hand, when one only has a circuit-based description of
the transition probabilities of a Markov chain over an
exponentially-large state space, there are complexity-theoretic
barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
%%Performance guarantees play a key role in the analysis and design of machine learning algorithms.
%
%In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
%Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$. Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
%distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
%$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
%The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
%\[
%  \tmix
%  :=
%  \min\Braces{
%    t \in \bbN :
%    \sup_{\vq}
%    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
%%    \Norm{
%%      \vq \vP^t - \vpi
%%    }_{\tv} 
%    \leq 1/4
%  }\,.
%\]
%Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
%to set $A$ by $\vpi$, and the supremum is over
%all possible initial distributions $\vq$.
%The problem studied in the paper is the construction of a confidence
%set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
%the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
%time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
%In the remainder of the introduction, we describe our motivation of studying this problem, followed by a 
%brief description of our main results.
%\todod{``might be interesting'' sounds weak.} \todoc{better now?}
%
%In numerous scientific applications and some machine learning tasks
%one is interested in estimating the mean $\vpi(f) = \sum_i \pi_i f(i)$ of some function $f$ of the states of a Markov chain. 
%For example, this is the realms of the celebrated Markov Chain Monte Carlo (MCMC) paradigm, but 
%the problem also arises in performance prediction when working with time correlated data, as is common in,
%for example, in reinforcement learning.
%%as 
%% (cf. \cref{sec:mot}) 
%Under our assumptions,
%the law of large numbers guarantees that the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$ 
%converges almost surely to $\vpi(f)$, while the central limit theorem tells us that as $n\to \infty$, 
%the distribution of the deviation $\sqrt{n}( \vpi_n(f)-\vpi(f))$ will be normal with zero mean and asymptotic
%variance $\sigma^2_{\mathrm{as}}(f) \defeq \lim_{n\to\infty} n \Var\Parens{ \vpi_n(f) }$. \todoc{shall we add a citation here?}
%While these results help us to understand how the sample mean over a Markov chain behaves in the limit,
%they tell us little about the finite time, non-asymptotic behavior of the sample mean,
%which are needed for prudent evaluation of one's method or even for algorithm design 
%\cite{
%MCMCDiscussion93%
%,DBLP:conf/valuetools/KontoyiannisLM06%
%,BBL06%
%,MniSzeAu08%
%,MauPo09%
%,LiLiWaSt11:KWIK%
%,flegal2011implementing%
%,Gyori-paulin15%
%,SwaJoa15:LoggedBandit%
%}. %BaHaVa10:activelearning,ShaVo08:conformalpred
%To address this need, numerous works have been devoted to deriving
%Chernoff-type, exponential bounds for upper bounding the tail probabilities $p(r) = \Pr\Parens{ |\vpi_n(f)-\vpi(f)|> r }$,
%also known as ``concentration bounds'', e.g., 
%\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
%Compared with the case when $(X_t)_t$ is an independent, identically distributed (iid) sequence of
%random variables, these tail probabilities must be larger due to the temporal dependence of the sequence $(X_t)_t$.
%Intuitively, for the Markov chain to forget the state where it is at time $t$, or in other words to get a 
%``fresh'' sample $X_{t'}$ that is approximately ``independent'' from $X_t$, one must wait $\Theta(\tmix)$ time steps.
%Thus, in the bounds on $p(r)$, one expects to see that the sample size $n$ 
%gets appropriately reduced as $\tmix$ grows and this is indeed the case in bounds mentioned above.
%is replaced by $n/\tmix$, which one
%might think of as an approximate sample size. 
%Indeed, all the above mentioned works get bounds on $p(r)$ that depend on various functions of $\vP$.
%Namely, Kontoyiannis et al. obtain bounds that depend on certain information-theoretic quantities defined implicitly
%as a function of $\vP$ and $f$, which are after simplification is bounded based on quantities related to 
%a regeneration-type quantity derived from $\vP$
%\cite{DBLP:conf/valuetools/KontoyiannisLM06}, while the other works mentioned above, in the case of reversible chains,
%derive bounds that depend on the spectral gap of $\vP$.
%Note that the spectral gap for such chains is essentially inversely related to the mixing time of the chain (see \cref{sec:setting}).
%%\cite{gillman1998chernoff,leon2004optimal,,paulin15}.
%Much research have been devoted to finding appropriate \emph{a priori} bounds on the various quantities that appear 
%in these bounds, mostly in the theory community (e.g., \cite{saloff2004random}). 
%To address the lack or looseness of such a priori bounds, a significant effort also went into developing methods
%that estimate these quantities based on data, mostly in the MCMC community in an effort to develop
%effective methods that are able to diagnose whether a chain has reached its stationary state 
%(e.g., \cite{MCMCDiscussion93,flegal2011implementing,Gyori-paulin15}).
%While the typical results state the asymptotic consistency of these estimators, or a central limit theorem (which 
%are of little help to achieve a fully empirical, non-asymptotic bound on $p(r)$),
%Gyori and Paulin give a finite time error bound for these estimates. 
%Unsurprisingly, these error bounds themselves
%depend on other unknown quantities of the Markov chain (which can be bounded in terms of the mixing time, again)
%\cite{Gyori-paulin15}.
%
%This situation is similar to what appears when one attempts to use a Bernstein-type deviation bound to 
%control the deviation between the sample mean of a sequence of iid random variables from their common mean
%with a known range, but unknown variance: For a tight bound, one wishes to use the unknown variance, but in the deviation
%bound for the error of a variance estimate, the variance appears again.
%While in the case of Bernstein-bounds this circular reference leads to a nontrivial bound
%\cite{audibert2009}, it appears that the situation is different in the Markov setting.
%Focusing on the spectral gap, we will see that if one attempts to use a similar technique,
%the set of feasible values based on the two error bounds always will never rule out that the spectral gap is zero,
%or that the mixing time is infinite. Can one even hope to obtain 
%after finitely many samples a nontrivial upper bound on the mixing time?
%Obviously, for any finite sample size, if the mixing time of the chain is larger than the sample size than a sound method
%must return infinity as the upper bound on the mixing time. Hence, the question is whether one can design a method
%that never fails and if the sample size is appropriately large, returns a nontrivial bound on the mixing time, adapting the time it needs to trap the mixing time in a fixed size confidence interval to the unknown mixing time.
%
%\hide{
%\subsection{Motivation}
%\label{sec:mot}
%One key competence in statistical learning is the ability to predict one's performance on future data.
%To make performance prediction possible, one starts by posing some assumptions on how the data arises.
%Simpler assumptions lead to simpler algorithms and analysis, while they may also limit applicability of the
%obtained results and algorithms.
%The most common assumption employed postulates that all data points, including past and future data points,
%are sampled independently of each other from a common underlying distribution (e.g., \cite{SSBD14:UnderstandingML}).
%This is helpful on two accounts: 
%The lack of dependence between the data points means that sample averages of bounded functions of the data 
%will converge at a uniform rate (irrespective of the distribution), while the assumption that all data points share the same
%distribution means that one can learn about the distribution of future data points based on past data.
%}
%
%\hide{
%It is well-known that contrary to the idealized assumptions of the standard
%learning model, real data is almost never iid.
%Fortunately, many classical results of statistics and statistical
%learning theory extend to the dependent data in a more or less
%seamless fashion, at least when the data is sufficiently ``mixing''.
%Such results include the central limit theorem, 
%exponential tail bounds for empirical processes, 
%performance guarantees for complexity regularization
%using sieves or roughness penalties, as well as 
%results based on stability arguments or Rademacher complexity 
%\citep{Woo86,Dou94,Yu94,ZheChu96,dlPenaGine99,Mei00,
%  MR1921877,vid2003,gamarnik03,LoKuScha05,MoRo08,MoRo09,
%  DBLP:conf/nips/SteinwartC09,Steinwart2009175,
%  FaSze11,
%  london:nips12asalsn,london:icml13,
%  DBLP:conf/nips/ShaliziK13}.
%
%Unfortunately, a distribution-free theory of non-iid learning is simply
%impossible, when arbitrary dependence among the sample points is allowed.
%Hence, one must quantify this dependence via the various mixing coefficients,
%which appear in the non-iid generalization bounds mentioned above.
%A frequently raised concern is that now the bounds depend on {\em
%unknown} properties of the distribution, which precludes their
%actual use in algorithmic applications (e.g., active
%learning~\citep{BBL06,DHM07}).
%%, and any algorithm aiming at a fully empirical error bound must
%%also estimate the mixing coefficients.
%In addition to learning-theoretic generalization bounds, the need for
%fully empirical confidence bounds arises in Markov Chain Monte Carlo
%(MCMC) simulations---so as to halt the run when sufficiently close to
%the stationary distribution \citep{galin06}, and in Markov Reward
%Processes \citep{sutton98}, as this will affect the learner's policy.
%As \citet{Mei00} notes, in the absence of accurate estimates of the
%mixing time, such fully empirical bounds remain ``a theoretician's
%dream''.
%
%Our goal here is to develop an algorithm that, given a single sample
%path of a Markov chain, computes non-trivial bounds on the mixing time
%of the Markov chain.
%We remark on the inherent chicken-and-egg nature of the challenge:
%confidence intervals derived from a Markov chain---say, on the the
%chain's mixing time---are typically given in terms of the chain's
%mixing time.
%Our way out of this conundrum is to give {\em empirical} (i.e.,
%\emph{data-dependent}) as opposed to {\em a priori} guarantees.
%To illustrate the intuition, consider a Markov chain on two states,
%$\braces{0,1}$.
%A learner who is stuck in state $0$ for a long time learns very little
%about the chain and hence cannot be expected to give meaningful upper
%bounds on its mixing time.
%Consider, however, a learner who is lucky enough
%to be in a chain that switches frequently between the two states.
%Surely at some finite time, the learner will possess meaningful and nontrivial
%information about the chain's mixing time --- crucially, without any
%{\em prior} knowledge regarding this quantity. This simple example captures
%the nature of our results: in some cases, we will not be able to say
%anything meaningful (but the learner will recognize that this is the case),
%while in ``luckier'' situations, we will be able to provide nontrivial
%upper estimates on the mixing time.
%}
%
%\paragraph{Main results.}
%We propose what is apparently the first efficient and fully empirical
%estimator of Markov mixing time.
%\todod{``apparently'' sounds weak.}
%Consider a reversible ergodic Markov chain on $d$ states with spectral
%gap $\gap$ (which is well-known to govern mixing time)
%and stationary distribution minorized by $\pimin$.
%Our main results are summarized as follows.
%\begin{enumerate}
%  \item
%    In \cref{thm:err}, we give an estimator for $\gap$ that
%    achieves multiplicative accuracy from \emph{a single sample path}
%    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
%    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
%    We also provide an estimator for $\pimin$ that requires a sample
%    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
%    However, the valid confidence intervals suggested by
%    \cref{thm:err} depend on the unknown quantities $\pimin$ and
%    $\gap$.
%
%  \item
%    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
%    (d\log d)/\gap + 1/\pimin }$ observations are necessary for
%    constant multiplicative accuracy in estimating $\gap$.
%
%  \item
%    In \cref{thm:empirical}, we develop a procedure for computing
%    \emph{valid empirical confidence intervals} for $\pimin$ and
%    $\gap$.
%    The procedure uses just a single sample path and does not require
%    any prior knowledge about $\pimin$ or $\gap$.
%    We also quantify the rate at which the width of the confidence
%    intervals shrink as a function of the length of the sample path.
%
%    We show in \cref{thm:combined} how to combine these empirical
%    confidence intervals with the non-empirical bounds from
%    \cref{thm:err} to produce valid empirical confidence intervals;
%    the width of these intervals converge to zero asymptotically as
%    fast as those from \cref{thm:err}.
%
%%    which is
%%    $     O\Parens{
%%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
%%      +
%%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
%%    }.$
%
%
%    
%%    \todoc{please add rate}
%
%\end{enumerate}
%
%
%
%\paragraph{Related work.} \todoc{Move this later? It also needs to be updated, to 
%adjust it to the current intro.}
%Owing to the wide applicability of MCMC methods to various computational
%problems, much effort has been devoted to developing effective convergence
%diagnostic methods the aim of deriving stopping rules or stationarity
%diagnoses. While a large number of heuristic techniques have been developed
%\citep[e.g.,][]{MCMCDiscussion93,CoCa96}, it is widely recognized that all
%these methods may fail to meet the guarantees that they are designed for.
%Some recent advance in this area \citep{1209.0703,Gyori-paulin15} have
%yielded practical algorithms for estimating the rates of convergence
%of various MCMC algorithms, but they fall short of providing intervals
%that are valid with any finite length sequence.
%
%In the theory community, \citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}
%developed algorithms for testing whether a given Markov chain over a
%finite state space of size $d$ is rapidly mixing.
%In particular, assuming access to a sampling oracle that can generate
%independent transitions from any given state, their algorithm can
%distinguish between chains that in $t$ steps for a fixed $t$ reach a
%distribution close to the stationary distribution using $\tilde{O}(t
%d^{5/3})$ oracle calls.
%It is remarkable that when $t$ is small enough, the procedure requires
%sublinearly many samples in the total number of transitions in the
%Markov chain.
%For Markov chains with exponentially-large state spaces (with a
%circuit-based description of the transition probability structure),
%\citet{BhaBoMo11} establishes complexity-theoretic barriers for
%similar MCMC diagnostic problems.
%%A similar question was studied by \citet{BhaBoMo11} in the context of
%%Markov chains with very large state spaces.
%%They assume an circuit based description of the
%%transition probability structure of the matrix and derive several hardness
%%results for Markov chains whose state space consists of all binary strings of
%%a given length. For example,
%%they prove that given a Markov chain that mixes
%%rapidly, it is $\mathsf{SZK}$-hard to distinguish whether the chain is close to
%%stationarity by time $t$ or far from stationarity at time $ct$ for a
%%constant $c$ when started from a given initial state ($\mathsf{SZK}
%%\subset \mathsf{AM} \cap \mathsf{coAM}$, which is conjectured to
%%extend beyond $\mathsf{BPP}$ \citep{GoVa11}).\footnote{%
%%  $\mathsf{SZK}$ stands for statistical zero-knowledge, $\mathsf{BPP}$
%%  stands for probabilistic polynomial-time, $\mathsf{AM}$ stands for
%%  ``Arthur-Merlin protocol''.%
%%}
%
%Other works include that of \citet{McDoShaSche11} who propose to estimate
%the $\beta$-mixing coefficients of a stationary process possessing a density
%based on histogram estimates of the underlying joint density. Their main
%result is a consistency result for a fixed ``lag''. For fast mixing processes
%(e.g., for $k$-step Markov processes) their result yields a convergence rate,
%which unfortunately is given in terms of the unknown mixing coefficients.
%%Catch-22 problem remains: The rate of convergence of their estimates depends on the unknown mixing coefficients
%%(Theorem 2.4)
%% $\exp(W(\log(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\log \log(n)$ and $\log n$).
%The earlier study of \citet{Nobel06} considers the generalization of
%existing discernibility results for iid processes to weakly uniformly
%ergodic families, and shows how this can be used to study---and in
%some cases estimate---polynomial mixing rates for dependent processes.
%
%

>>>>

This work
tackles the challenge of constructing
fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The \emph{mixing time} $\tmix$ of the Markov chain is the
number of time steps required
for the chain to
be within a fixed threshold of
its stationary
distribution:
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that
succeeds with probability $1-\delta$
in trapping
the value of the
mixing time $\tmix$.

This problem is motivated by the numerous scientific applications and
machine learning tasks in which
the quantity of interest is
the
mean $\vpi(f) = \sum_i \pi_i f(i)$
for some function $f$ of the states of
a Markov chain.
This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
paradigm~\cite{liu2001monte}, but the problem also arises in
performance prediction involving time-correlated data, as is common in
reinforcement learning~\cite{sutton98}.
Observable bounds on mixing times are useful in the design and
diagnostics of these methods; they yield effective approaches to
assessing the estimation quality, even when \emph{a priori} knowledge
of the mixing time or correlation structure is unavailable.

\paragraph{Main results.}
We develop the first procedure for constructing non-trivial
and fully empirical confidence intervals for Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with absolute
spectral gap $\gap$ and stationary distribution minorized by $\pimin$. 
As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
\begin{equation}
  \label{eq:mixing-time-bound}
  \Parens{\mfrac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \mfrac1{\gap} \ln \mfrac4{\pimin}
  ,
\end{equation}
and hence it suffices to estimate $\gap$ and $\pimin$. 
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{sec:rates-lower}, we show that in some problems 
    $n = \Omega\parens{ (d\log d)/\gap + 1/\pimin}$ observations are necessary for any procedure
    to guarantee constant multiplicative accuracy in estimating $\gap$
    (\cref{thm:lb-pimin,thm:lb-gap}).
    Essentially, in some problems \emph{every} state may need to be visited about
    $\log(d)/\gap$ times, on average, before an accurate estimate of
    the mixing time can be provided, regardless of the estimation
    procedure used.

  \item
    In \cref{sec:rates-upper}, we give a point-estimator for $\gap$,
    and prove in \cref{thm:err} that it achieves multiplicative
    accuracy from \emph{a single sample path} of length
    $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide a point-estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    This establishes the feasibility of \emph{estimating} the mixing
    time in this setting.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.
    We also discuss the importance of reversibility, and some possible
    extensions to non-reversible chains.

  \item
    In \cref{sec:empirical}, the construction of
    valid \emph{fully empirical confidence intervals} for $\pimin$ and
    $\gap$ are considered. First, the difficulty of the task is explained,
    i.e., why the standard approach of turning the finite time confidence intervals of
    \cref{thm:err} into a fully empirical one fails.
	Combining several results from perturbation theory in a novel fashion
	we propose a new procedure and prove that it avoids slow convergence
    (\cref{thm:empirical}).
    We also explain how to combine the empirical confidence intervals
    from \cref{alg:empest} with the non-empirical bounds from
    \cref{thm:err} to produce valid empirical confidence intervals.
    We prove in \cref{thm:combined} that the width of these new
    intervals converge to zero asymptotically at least as fast as
    those from either \cref{thm:err} and \cref{thm:empirical}.

@@

Although these asymptotic results help us understand the limiting
behavior of the sample mean over a Markov chain, they say little about
the finite-time non-asymptotic behavior, which is often needed for the
prudent evaluation of a method or even its algorithmic design
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}.
To address this need, numerous works have developed Chernoff-type
bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
valuable tools for non-asymptotic probabilistic
analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,
paulin15}.
These probability bounds are larger than corresponding bounds for
independent and identically distributed (iid) data due to the temporal
dependence; intuitively, for the Markov chain to yield a fresh draw
$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
$\Theta(\tmix)$ time steps.
Note that the bounds generally depend on distribution-specific
properties of the Markov chain (e.g., $\vP$, $\tmix$, $\gap$), which are often
unknown \emph{a priori} in practice.
Consequently, much effort has been put towards estimating these
unknown quantities, especially in the context of MCMC diagnostics, in
order to provide data-dependent assessments of estimation
accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
However, these approaches generally only provide asymptotic
guarantees, and hence fall short of our goal of empirical bounds that
are valid with any finite-length sample path.

@@

It is possible to eliminate many of the difficulties presented above
when allowed more flexible access
to the Markov chain.
For example, given a sampling oracle that generates
independent transitions from any given state (akin to a ``reset''
device), the mixing time becomes an efficiently testable property in
the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
On the other hand, when one only has a circuit-based description of
the transition probabilities of a Markov chain over an
exponentially-large state space, there are complexity-theoretic
barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
